---
title: "Reproducible Research: Peer Assessment 1"
author: "Zaixing Shi, 07/15/2015"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First thing first, load the data
```{r}
act <- read.csv("activity.csv",header=T)
```

Now let's check the data

```{r}
str(act)
head(act,10)
```


Note that variable date was not a date vector, so let's convert it into date 
format
```{r}
act$date <- as.Date(as.character(act$date))
```

Check to see it's current format
```{r}
class(act$date)
```


## What is mean total number of steps taken per day?

Take a look at the distribution of total number of steps taken each day
```{r}
hist(tapply(act$steps,act$date,sum,na.rm=T),xlab="total number of steps",
     main="Histogram of daily total steps")
```

To get a better sense of the distribution of daily steps, let's look at the 
mean and median of daily total steps
```{r}
c(Mean=mean(tapply(act$steps,act$date,sum,na.rm=T)),
  Median=median(tapply(act$steps,act$date,sum,na.rm=T)))
```

So on average the sample walked 9354 steps each day, and half of the sample
walked more than 10395 steps daily. Who are these energetic people? 


## What is the average daily activity pattern?

To answer this question, we should get the average steps across all dates
```{r}
mean_int_steps <- tapply(act$steps,act$interval,mean,na.rm=T)
```
Then the average number of steps during every 5-minute interval were like this:

```{r}
plot(x=names(mean_int_steps),y=mean_int_steps,type="l",
     xlab="Intervals",ylab="Average steps",
     main="Average steps for each interval")
```

In this plot, the Empire State of Steps occured at the interval of... 
```{r}
names(mean_int_steps[mean_int_steps==max(mean_int_steps)])
```

That is, on average our sample walked the most during the five-minute interval between 8:35-8:40 -- probably the morning rush hour? 

## Imputing missing values
To learn how many missing values were present:
```{r}
nrow(act[which(is.na(act$steps)),])
```

So 2304 intervals in our data had missing steps. A huge problem!

To make educated guesses about these missing steps, I make this bold statement
that "All missing steps were created equal"" -- that is, all steps at a given 
interval should be the same.  Therefore we will fill in the missing steps using
the average step at that particiular interval.

```{r}
act.fill <- act
act.fill$steps[is.na(act.fill$steps)] <- lapply(names(mean_int_steps),function(x) 
 act$step[act.fill$interval==x & is.na(act.fill$step)] <- 
   mean_int_steps[names(mean_int_steps)==x])
```

The dataset act.fill is the clone of dataset act but with missing steps filled 
in. To see if the fill in was successful:
```{r}
head(act)
head(act.fill)

```

To make the histogram of daily total steps, let's round up the filled in steps
and then plot.

```{r}
act.fill$steps <- round(as.numeric(act.fill$steps))
hist(tapply(act.fill$steps,act.fill$date,sum,na.rm=T),
     xlab="total number of steps",
     main="Histogram of daily total steps")

```

And here are the comparisons between our means and medians before and after
filling in missing steps:
```{r}
c(Mean.before=mean(tapply(act$steps,act$date,sum,na.rm=T)),
  Mean.after=mean(tapply(act.fill$steps,act.fill$date,sum,na.rm=T)),
  Median.before=median(tapply(act$steps,act$date,sum,na.rm=T)),
  Median.after=median(tapply(act.fill$steps,act.fill$date,sum,na.rm=T)))
```

So apparently, filling in the missing steps will increase the mean and median,
which makes perfect sense as in pervious calculation we excluded missing values.

To make formal comparisons between the original and imputed steps, let's do
a t-test of the means:
```{r}
t.test(tapply(act$steps,act$date,sum,na.rm=T),
       tapply(act.fill$steps,act.fill$date,sum,na.rm=T))
```

So the imputation increased the mean, but not caused significant increase to 
the daily total steps (p-value for t-test = 0.1032).

## Are there differences in activity patterns between weekdays and weekends?
Create a factor variable indicating weekday or weekend:
```{r}
act$weekdays <- ifelse(weekdays(act$date) %in% c("Saturday","Sunday"),
                       "Weekend","Weekday")
```
To check if we correctly coded the weekday/weekend:
```{r}
table(weekdays(act$date),act$weekdays)
```
Looks cool!

Now let's plot the time series by weekday and weekend.
```{r}
invisible(lapply(c("Weekday","Weekend"),function(x) {
  mean_steps <- tapply(act[which(act$weekdays==x),"steps"],
                       act[which(act$weekdays==x),"interval"],mean,na.rm=T)
    plot(x=names(mean_steps),y=mean_steps,type="l",ylim=c(0,300),
     xlab="Intervals",ylab="Average steps",
     main=paste("Average Steps for Intervals on a ",x,sep=""))
  
  }
  ))
```

So it appears that our sample tend to be more active during the weekend - instead
of the bi-modal walking pattern observed during weekedays, sample walked more
evenly and more frequently during weekends.  

We can further compare the mean steps for each intervals between weekdays and
weekend.

```{r}
t.test(tapply(act[which(act$weekdays=="Weekday"),"steps"],
                       act[which(act$weekdays=="Weekday"),"interval"],
              mean,na.rm=T),
       tapply(act[which(act$weekdays=="Weekend"),"steps"],
                       act[which(act$weekdays=="Weekend"),"interval"],
              mean,na.rm=T),
       paired=T)


```

Compared with their performance during weekdays, our sample walked about 8 more steps in every 5-minute intervals during weekend. 
And this difference is statistically significant (P=0.001853).

I'm so glad they had a chance to relax after a whole week of work!














